A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-07 inst/manuscript/manuscript.Rmd
section 3,4,5 updates

FBPK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean zero and a spatial correlation structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative (between zero and one), second-order stationary (depending only on the distance between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$, though @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or both. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (for a thorough list of spatial covariance functions, see @cressie1993statistics. The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter quantifying the variability that is dependent (coarse-scale), $\sigma^2_{2}$ is the variance parameter quantifying the variability that is independent (fine-scale), $\phi$ is the range parameter measuring the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. The proportion of variability attributable to dependent random error is $\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})$. Similarly, the proportion of variability attributable to independent random error is $\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})$. Finally we note that $\sigma^2_{1}$ and $\sigma^2_{2}$ are often called the partial sill and nugget, respectively.

@@

# Numerical Study {#sec:numstudy}

We used a simulation study to investigate performance of four sampling-analysis combinations: IRS-Design, IRS with a design-based analysis; IRS-Model, IRS with a model-based analysis; GRTS-Design, GRTS sampling with a design-based analysis; and GRTS-Model, GRTS sampling with a model-based analysis. These combinations are also provided in Table \ref{tab:designanalysis}.

@@

Performance of the four sampling-analysis combinations was evaluated in 36 different simulation scenarios. The 36 scenarios resulted from the crossing of three sample sizes, two location layouts, two response types, and three proportions of dependent random error. The three sample sizes ($n$) were $n = 50, n = 100,$ and $n = 200$. Samples were always selected from a population size ($N$) of $N = 900$. The two location layouts were random and gridded. Locations in the random layout were selected randomly from the unit square ($[0, 1] \times [0, 1]$). Locations in the gridded layout were selected randomly on a fixed grid from the unit square. The two response types were normal and lognormal. For the normal response type, the response was simulated using mean-zero random errors with the exponential covariance (Equation$~$\ref{equation:expcov}) for varying proportions of dependent random error. The proportion of dependent random error is represented by $\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)$, where $\sigma^2_1$ and $\sigma^2_2$ are from Equation$~$\ref{equation:expcov}. The total variance, $\sigma^2_1 + \sigma^2_2$, was always 2. The range was always $\sqrt{2} / 3$, which means that the correlation in the dependent random error decayed to nearly zero at the largest possible distance between two units in the domain. For the lognormal response type, the response was first simulated using the same approach as for the normal response type, except that the total variance was 0.6931 instead of 2. The response was then exponentiated, yielding a random variable whose total variance is 2. The lognormal responses were used to evaluate performance of the sampling-analysis approaches for data that were skewed.

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. In each trial, IRS and GRTS samples were selected and then design-based and model-based analyses were used to estimate the mean and construct confidence (design-based) or prediction (model-based) intervals. We recorded the bias, squared error, and interval coverage for all sampling-analysis combinations in each trial. Then we summarized the performance of the combinations across trials by calculating average bias, rMS(P)E (root-mean-squared error for the design-based approaches and root-mean-squared-prediction error for the model-based approaches), and the rate at which the true mean is contained in its 95% interval. The GRTS algorithm and the local neighborhood variance estimator are available in the \textbf{\textsf{R}} package `spsurvey` [@dumelle2021spsurvey]. FPBK is available in the `sptotal` \textbf{\textsf{R}} package [@higham2021sptotal] and covariance parameters were estimated using Restricted Maximum Likelihood  [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing].

@@

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four approaches from Table \ref{tab:designanalysis} using the random location layout with "IRS-Design" as the baseline. More formally, the relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}
When there is no spatial correlation (Figure \ref{fig:figeff}, top row), the four sampling-analysis combinations have approximately equal rMS(P)E. So, using GRTS or using a spatial model does not result in much, if any, loss in efficiency even when the response variable is not spatially correlated. When there is spatial correlation (Figure \ref{fig:figeff}, middle and bottom row), the GRTS-Model combination tends to perform best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though the difference in relative rMS(P)E among IRS-Model, GRTS-Design, and GRTS-Model is relatively small.  As the strength of spatial correlation increases, the gap in rMS(P)E between IRS-Design and the other combinations widens. Finally we note that when there is spatial correlation, IRS-Model outperforms IRS-Design by a large margin, suggesting that the poor design properties of IRS are largely mitigated by the model-based analysis. These conclusions are similar to those observed in the grid location layout. Tables for rMS(P)E in all 36 simulation scenarios are provided in the supplementary material.

@@

We also studied 95% interval coverage among the combinations. The design-based 95% confidence intervals and model-based 95% prediction intervals were constructed using the normal distribution. Justification for the design-based and model-based intervals comes from the asymptotic normality of totals via the Central Limit Theorem.

Figure \ref{fig:figconf} shows the 95% interval coverage for each of the four combinations in the random location layout. All four combinations have fairly similar interval coverage within each scenario. Coverage in the normal response scenarios tended to be near 95% and slightly higher than coverage in the lognormal scenarios. Coverage in the lognormal scenarios still generally exceeded 90%. Coverage tended to always increase with the sample size. At a sample size of 200, all four combinations had approximately 95% interval coverage in both response scenarios and all dependent error proportions.  These conclusions were similar to those found in the grid location layout.
Tables for interval coverage in all 36 simulation scenarios are provided in the supplementary material.

@@

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and water quality in lakes in the contiguous United States, to study mercury concentration. Although we know the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes.

@@

We selected a single IRS sample and a single GRTS sample and estimated the mean mercury concentration and its standard error using using design-based and model-based approaches; Table \ref{tab:appliedtab} shows the results. For all four sampling-analysis combinations, the true realized mean mercury concentration is within the bounds of the 95% intervals. However, we should not generalize these results to any other data or even to other samples from these data. But, we do note a couple of patterns. The design-based IRS analysis shows the largest standard error: a likely reason is that this is the only approach that does not incorporate any spatial information regarding mercury concentration across the contiguous United States. We also see that both approaches using the GRTS sample have a lower standard error than the both approaches using the IRS sample. We would expect this to be the case for most samples because mercury concentration exhibits spatial patterning, so a spatially balanced sample should usually yield a lower standard error.

To better understand the dependence structure in mercury concentration, the empirical semivariogram and corresponding fit of the model-based approaches can be visualized. The empirical semivariogram quantifies the halved squared differences (semivariance) among response values at different distances apart. If a process exhibits strong spatial dependence, the empirical semivariogram will have small values at small distances and large values at large distances. Figure \ref{fig:figsv} shows the empirical semivariogram for GRTS-Model, displaying the average semivariance for several distances.  Overlain onto Figure \ref{fig:figsv} is the estimated semivariance obtained using the covariance parmaeters from the REML fit of GRTS-Model. Figure \ref{fig:figsv} provides evidence that there is strong correlation in mercury concentration among the sites.

@@

The design-based and model-based approaches to inference are fundamentally different paradigms by which samples are selected and data are analyzed. The design-based approach incorporates randomness through sampling to estimate a population parameter. The model-based approach incorporates randomness through distributional assumptions to predict the realized values of a random process. Though these approaches have often been compared in the literature both from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling. Aside from the theoretical differences described, a few analytical findings from the simulation study are particularly notable. First, the sampling decision (GRTS vs IRS) is most important when using a design-based analysis. Though GRTS-Model still outperformed IRS-Model, the model-based analysis mitigated much of the inefficiency of the IRS sample. Second, independent of the analysis approach, there is no reason to use IRS over GRTS for sampling spatial data, as GRTS-Design and GRTS-Model generally performed at least as well as their IRS counterparts when there was no spatial correlation and noticeably better than there IRS counterparts when there was spatial correlation. Third, The stronger the spatial correlation, the larger the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations. Fourth and finally, interval coverage for the normal response was very close to 95% for all sample sizes, while interval coverage for the lognormal response was not very close to 95% until $n = 200$.

There are several benefits and drawbacks of the design-based and model-based approaches for spatial data, some of which we have not yet discussed but are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based estimation of covariance parameters can be computationally burdensome, especially for likelihood-based methods such as REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression formulation commonly used to handle binary data in a model-based approach. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and the response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the process of study. Model selection is also possible using model-based approaches and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Model-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. The benefits and drawbacks of both approaches, alongside our theoretical and analytical comparisons, should be seriously considered when choosing among them. This is especially true from an analysis perspective, as we found that using a spatially balanced sampling algorithm benefits both design-based and model-based analyses.

>>>>

FBPK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean-zero and a spatial correlation structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative (between zero and one), second-order stationary (depending only on the distance between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$, though @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or both. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (for a thorough list of spatial covariance functions, see @cressie1993statistics. The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter quantifying the variability that is dependent (coarse-scale), $\sigma^2_{2}$ is the variance parameter quantifying the variability that is independent (fine-scale), $\phi$ is the range parameter measuring the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. The proportion of variability attributable to dependent random error is $\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})$. Similarly, the proportion of variability attributable to independent random error is $\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})$. Finally we note that $\sigma^2_{1}$ and $\sigma^2_{2}$ are often called the partial sill and nugget, respectively.

@@

# Simulation Study {#sec:numstudy}

We used a simulation study to investigate performance of four sampling-analysis combinations: IRS with a design-based analysis, called "IRS-Design"; IRS with a model-based analysis, called "IRS-Model"; GRTS sampling with a design-based analysis, called "GRTS-Design"; GRTS sampling with a model-based analysis, called "GRTS-Model". These combinations are also provided in Table \ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in 36 different simulation scenarios. The 36 scenarios resulted from the crossing of three sample sizes, two location layouts, two response types, and three proportions of dependent random error. The three sample sizes ($n$) were $n = 50, n = 100,$ and $n = 200$. Samples were always selected from a population size ($N$) of $N = 900$. The two location layouts (of the population units) were random and gridded. Locations in the random layout were randomly generated inside the unit square ($[0, 1] \times [0, 1]$). Locations in the gridded layout were placed on a fixed, equall spaced grid inside the unit square. The two response types were normal and lognormal. For the normal response type, the response was simulated using mean-zero random errors with the exponential covariance (Equation$~$\ref{equation:expcov}) for varying proportions of dependent random error. The proportion of dependent random error is represented by $\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)$, where $\sigma^2_1$ and $\sigma^2_2$ are the dependent random error variance (partial sill) and independent random error variance (nugget), respectively, from Equation$~$\ref{equation:expcov}. The total variance, $\sigma^2_1 + \sigma^2_2$, was always 2. The range was always $\sqrt{2} / 3$, which means that the correlation in the dependent random error decayed to nearly zero at the largest possible distance between two units in the domain. For the lognormal response type, the response was first simulated using the same approach as for the normal response type, except that the total variance was 0.6931 instead of 2. The response was then exponentiated, yielding a random variable whose total variance is 2. The lognormal responses were used to evaluate performance of the sampling-analysis approaches for data that were skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. In each trial, IRS and GRTS samples were selected and then design-based and model-based analyses were used to estimate (design-based) or predict (model-based) the mean and construct confidence (design-based) or prediction (model-based) intervals. Then we recorded the bias, squared error, and interval coverage for all sampling-analysis combinations. After all 2000 trials, we summarized the long-run performance of the combinations by calculating average bias, rMS(P)E (root-mean-squared error for the design-based approaches and root-mean-squared-prediction error for the model-based approaches), and the proportion of tiems the true mean is contained in its 95% interval. The GRTS algorithm and the local neighborhood variance estimator are available in the \textbf{\textsf{R}} package `spsurvey` [@dumelle2021spsurvey]. FPBK is available in the `sptotal` \textbf{\textsf{R}} package [@higham2021sptotal] and covariance parameters were estimated using Restricted Maximum Likelihood  [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing].

@@

Figure \ref{fig:figeff} shows the relative rMS(P)E of the four approaches from Table \ref{tab:designanalysis} using the random location layout with "IRS-Design" as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}
When there is no spatial correlation (Figure \ref{fig:figeff}, "Prop DE: 0" row), the four sampling-analysis combinations have approximately equal rMS(P)E. So using the GRTS sampling plan or a model-based analysis does not result in much, if any, loss in efficiency compared to IRS-Design when there is no spatial correlation. When there is spatial correlation (Figure \ref{fig:figeff}, "Prop DE: 0.5" and "Prop DE: 0.9" rows), GRTS-Model tends to perform best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and IRS-Model is relatively small.  As the strength of spatial correlation increases, the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations widens. Finally we note that when there is spatial correlation, IRS-Model outperforms IRS-Design by a large margin, suggesting that the poor design properties of IRS are largely mitigated by the model-based analysis. These conclusions are similar to those observed in the grid location layout, so we omit a grid location layout figure here. Tables for rMS(P)E in all 36 simulation scenarios are provided in the supplementary material.

@@

We also studied 95% interval coverage among the sampling-analysis combinations. The design-based confidence intervals and model-based prediction intervals were constructed using the normal distribution. Justification for this comes from the asymptotic normality of means via the Central Limit Theorem.

Figure \ref{fig:figconf} shows the 95% interval coverage for each of the four sampling-analysis combinations in the random location layout. Within each scenario, the sampling-analysis combinations tend to have fairly similar interval coverage. Coverage in the normal response scenarios was usually near 95%, while coverage in the lognormal response scenarios varied from from 90% to 95%. Coverage tended to always increase with the sample size. At a sample size of 200, all four sampling-analysis combinations had approximately 95% interval coverage in both response scenarios for all dependent error proportions.  These conclusions are similar to those observed in the grid location layout, so we omit a grid location layout figure here. Tables for interval coverage in all 36 simulation scenarios are provided in the supplementary material.

@@

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and water quality for lakes in the contiguous United States. Specifically, we will analyze mercury concentration in lakes. Although we know the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes.

@@

We selected a single IRS sample and a single GRTS sample and estimated (design-based) or predicted (model-based) the mean mercury concentration and its standard error using using design-based and model-based approaches. For the model-based analyses, the exponential covariance was used. Table \ref{tab:appliedtab} shows the results from these analyses. For all four sampling-analysis combinations, the true realized mean mercury concentration is within the bounds of the 95% confidence (design-based) or prediction (model-based) intervals. Though we should not generalize these results to other samples from these data, we do note a couple of patterns. The design-based IRS analysis shows the largest standard error: a likely reason is that this is the only approach that does not incorporate any spatial information regarding mercury concentration. Both analyses using GRTS sampling have lower standard errors than both analyses using IRS sampling. We expect that these patterns are consistent with other samples from these data because mercury concentration exhibits spatial patterning, so a spatially balanced sample should usually yield a lower standard error.

To better understand the dependence structure in mercury concentration, the empirical semivariogram can be visualized alongside a fit using the model-based approaches. The empirical semivariogram quantifies the halved squared differences (semivariance) among response values at different distances apart. If a process exhibits strong spatial dependence, the empirical semivariogram will have small values at small distances and large values at large distances. Figure \ref{fig:figsv} shows the empirical semivariogram quantifying the average semivariance for several distances.  Overlain onto Figure \ref{fig:figsv} is the estimated semivariance obtained using the covariance parmaeters from the REML fit of GRTS-Model. Figure \ref{fig:figsv} provides evidence that there is strong correlation in mercury concentration among the sites.

@@

The design-based and model-based approaches to statistical inference are fundamentally different paradigms by which samples are selected and data are analyzed. The design-based approach incorporates randomness through sampling to estimate population parameters. The model-based approach incorporates randomness through distributional assumptions to predict realized values of a random process. Though these approaches have often been compared in the literature both from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling. Aside from the theoretical differences described, a few analytical findings from the simulation study are particularly notable. First, the sampling decision (GRTS vs IRS) is most important when using a design-based analysis. Though GRTS-Model still outperformed IRS-Model, the model-based analysis mitigated much of the inefficiency of the IRS sample. Second, independent of the analysis approach, we found no reason to prefer IRS over GRTS for sampling spatial data -- GRTS-Design and GRTS-Model generally performed at least as well as their IRS counterparts when there was no spatial correlation and noticeably better than their IRS counterparts when there was spatial correlation. Third, as the strength of spatial correlation increases, the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations also increases. Fourth and finally, when the response was normal, interval coverage for all sampling-analysis combinations was very close to 95% for all sample sizes; when the response was lognormal, interval coverage for all sampling and analysis was between 90% and 95% and closest to 95% when $n = 200$.

There are several benefits and drawbacks of the design-based and model-based approaches for finite population spatial data. Some we have discuss, but others we have not and they are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based estimation can be computationally burdensome, especially for likelihood-based methods such as REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression framework commonly used to analyze binary data in a model-based approach. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the dependence structure in the process of study. Model selection is also possible using model-based approaches and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Model-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. The benefits and drawbacks of both approaches, alongside our theoretical and analytical comparisons, can motive the process of choosing among them. This is especially true from an analysis perspective, as we found that using a spatially balanced sampling algorithm benefits both design-based and model-based analyses.
